The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-31 paper.qmd
moved the last section around a bit

@@

## The bivariate map
@fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a good uncertainty visualisation? And if not, why?

@@

## What makes uncertainty visualisation difficult?
Uncertainty visualisation is a uniquely difficult problem. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty. This difficulty is often described as a high-dimensional visualisation problem, as the difficulty in uncertainty visualisation is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006]. The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable, but we are not interested in visualising uncertainty as though it were any other variable.

@@

## What even is "uncertainty" anyway?
The pixel map is not the best uncertainty visualisation, but it is trying to *replicate* the best uncertainty visualisation. The best uncertainty visualisation is probably the the raw data, to understand why, we need to understand what "uncertainty" actually is.

Uncertainty is not just variance, and

@@

@@

## Current methods
### Secondary effects

- trust, confidence, etc

### Direct questions
Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].

@@

# ------------------- BELOW HERE IS UNSORTED ------------------------

### What is uncertainty?
There does not seem to be a clear agreed upon definition of uncertainty which would obviously create some issues when we want to visualize it. You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.

The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies. In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.

What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data. That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated. The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition. 
Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.

It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023]. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e. that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023]. This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.

This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with. Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate. Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake. We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty. The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves. This confusion is common and there are many papers that spend a great deal of time clarifying the difference. @Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances. @Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.

 Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation. Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community. Some authors simply confuse varianc for uncertainty. For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.  @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty. It is likely impossible to suppress all possible signals at once.

# How do we evaluate uncertainty visualisations?

## Current attempts to measure uncertainty
There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].

@@

## How to test uncertainty as noise
So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation. In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying. Unlike statistical models, visualisations are used in both data exploration and communication. This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is. Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation. Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics. These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature. This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.

@@

## Other elements of uncertainty visualisations
- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty

### Heuristics
Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].

##
In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations". Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations. Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.

>>>>

## What is uncertainty

@@

## The bivariate map
@fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is technically* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic did not hide the invalid signal, so it is not performing signal-suppression as we would like. At this point, it might be reasonable to ask, why? Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?

@@

## Why this approach may (or may not) work
Unfortunately it is not enough to only understand the purpose of uncertainty visualisation. To do signal-suppression successfully we need to understand what uncertainty actually *is*, as this will inform us on how it should be integrated into the visualisaiton.

### It's a variable... it's a metadata... it's uncertainty?
Describing what uncertainty *is* is actually suprisingly hard. Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty. Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.

There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. Rather than trying to define uncertainty by what it *is* it may be easier to try and describe what uncertainty *is not*. Thankfully, the literature on what uncertainty is *not* is much more specific.

Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", i.e. we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. Positivists abore the assumption of the "uniformity of nature" as they believed that reference to the unobservable is not true science. In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the fact inferential statistics introduces uncertainty.

This history lesson illustrates what uncertainty actually is. A by-product of inference. This means that anything that could lead us to violate our uniformity of nature assumption, is a source of uncertainty. The qualitative characteristics of uncertainty provided by our taxonomies all highlight aspect

feared 
This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty. 
While we may agree on the *purpose* of uncertainty, it is an entirely different can of worms to agree on what uncertainty actually *is*. @fig-bivariate uses variance as a proxy for uncertainty, however those terms are not perfect synonyms.

### Visualising the "single integrated uncertain value"
Uncertainty visualisation is a uniquely difficult problem. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty. This difficulty is often described as a high-dimensional visualisation problem, as the difficulty in uncertainty visualisation is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006]. The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable, but we are not interested in visualising uncertainty as though it were any other variable.

@@

## What even is "uncertainty" anyway?
The pixel map is not the best uncertainty visualisation, but it is trying to *replicate* the best uncertainty visualisation. The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.

## variance is not uncertainty
This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with. Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate. Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake. We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty. The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves. This confusion is common and there are many papers that spend a great deal of time clarifying the difference. @Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances. @Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.

You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.

### Goal/task dependence in each stage
Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].

### Problems with uncertainty having no definition
The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

### EDA is descriptive statistics
 Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation. Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community. Some authors simply confuse varianc for uncertainty. For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.  @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty. It is likely impossible to suppress all possible signals at once.

@@

If you have uncertainty information, the best approach to create an "uncertainty visualisation" might be to incorperate that information into the visualisation to attempt to replicate the original sample.

@@

@@

There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].

@@

### Heuristics
- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty

Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].

In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations". Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations. Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.

So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation. In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying. Unlike statistical models, visualisations are used in both data exploration and communication. This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is. Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation. Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics. These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature. This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.

@@
